Grand Prix motorcycle racing

### 2021-01-29 inst/moto.Rmd
consistentification

---
title: "Grand Prix motorcycle racing"
output: html_document
---

@@

Here I analyse a series of Grand Prix motorcycling race results using
the `hyper` package which implements the Plackett-Luce likelihood
function.  Consider the following dataset, taken from Wikipedia:

```{r cars}
data(motoGP)
motoGP_2019_table
```

Each row is a driver (rider?) and each column (after the first) a
venue.  Takin the first row we see that Marquez came second in Qatar,
first in Argentina, retired in Austin, and so on.  In the first
column, we see the result from Qatar (QAT) in which Marquez came
second, Dovizioso first, Vinales third, and so on.  Notation used also
includes six different classes of no-score such as "Ret" for retired,
"WD" for withdrawn, and so on.  I now use bespoke R function
\code{ordertable2supp()}, that translates data of this type into a
log-likelihood.  The function is similar to, but slightly different
from, the analysis in `eurovision.R`.

```{r,applydata}
head(motoGP_2019)
```

(object `motoGP_2019` is equivalent to
`ordertable2supp(motoGP_2019_table)`.  Above, we see a small part of
the Placket-Luce likelihood function for the 2019 results.  We can
find the maximum likelihood estimator by using the `maxp()` function:

```{r,maxp2017}
options(width = 100)
m <- maxp(motoGP_2019)
m
dotchart(m,pch=16)
pie(m)
```

In the above, note the dominance of Marquez.  It is sometimes
preferable to plot the estimated strengths on a log scale:

```{r,maxp2017log}
dotchart(log(m[1:27]),pch=16)
```

(in the diagram above, we remove the bottom-ranked player, Lecuona,
whose estimated strength is zero).

# Likelihood scoring vs points scoring

The championship winning order is determined by a points system: the
winner is awarded 25 points, second place 18 points, and so on.  We
can now check to see how the points total matches up with the
likelihood estimation.

```{r,pointsforwinning}
(points_awarded <- c(`1st`=25, `2nd`=20, `3rd`=16, `4th`=13, `5th`= 11,
`6th`=10, `7th`= 9, `8th`=8,`9th`=7, `10th`=6, `11th`=5, `12th`=4,
`13th`=3, `14th`=2, `15th`=1))
```

```{r,likelihoodandscore}
points <- motoGP_2019_table$points
names(points) <- motoGP_2019_table$driver

ox <- rank(-points) # negative because number 1 = largest number of points
oy <- rank(-m)      # negative because number 1 = largest strength
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,28),ylim=c(0,28),pch=16,
xlab="official order",ylab="my order",main='motoGP, 2019 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

In the graph above, we see a certain amount of disagreement.  Only the
top two riders, Marquez and Dovizioso, have rankings common to
likelihood system and points system.  Considering the next four
players (Vinales, Rins, Quartaro, Petrucci) we see that the
points-based order is Vinales, Rins, Petrucci, but the likelihood
order is Petrucci, Vinales, Rins (Rossi), Quartararo.

>>>>

---
title: "Grand Prix motorcycle racing"
output: bookdown::html_document2
---

@@

Here I analyse a series of Grand Prix motorcycling race results using
the `hyper2` package which implements the Plackett-Luce likelihood
function.  Consider the following dataset, taken from Wikipedia:

```{r cars}
moto_table <- read.table("motoGP_2019.txt")
head(moto_table)
```

Each row is a driver (rider?) and each column (after the first) a
venue.  Taking the first row we see that Marquez came second in Qatar
(QAT), first in Argentina, retired in Austin, and so on.  In the first
column, we see the result from Qatar (QAT) in which Marquez came
second, Dovizioso first, Vinales third, and so on.  Notation used also
includes six different classes of no-score such as "Ret" for retired,
"WD" for withdrawn, and so on.  I now use package function
`ordertable2supp()`, that translates data of this type into a
log-likelihood.  In the package, object `moto` is the corresponding
Plackett-Luce likelihood function; we drop the final column which is
the points scored by each rider.

```{r,showmotolikelihoodhead}
moto <- ordertable2supp(moto_table[,-ncol(moto_table)])
head(moto)
```

Above, we see a small part of the Placket-Luce likelihood function for
the 2019 results.  We can find the maximum likelihood estimator by
using the `maxp()` function:

```{r,maxpcalculator,cache=TRUE}
moto_maxp <- maxp(moto)
moto_maxp
```

```{r dotchartmoto, echo=TRUE,fig.cap="Dot chart of riders' strengths"}
dotchart(moto_maxp,pch=16)
```

```{r piechartmoto,echo=TRUE,fig.cap="Pie chart of riders' strengths"}
pie(moto_maxp)
```

In figures \@ref(fig:dotchartmoto) and \@ref(fig:piechartmoto), note
the dominance of Marquez.  It is sometimes preferable to plot the
estimated strengths on a log scale:

```{r dotchartlogmoto, echo=TRUE,fig.cap="Dot chart of riders' strengths, log scale"}
dotchart(log10(moto_maxp),pch=16)
```

(in figure \@ref(fig:dotchartlogmot) observe that the bottom-ranked
player, Lecuona, has a ML strength of zero but the estimate is
slightly positive for numerical reasons).

### Package dataset {-}

Following lines create `moto.rda`, residing in the `data/` directory
of the package.

```{r,label=savemotodataset}
save(moto,moto_table,moto_maxp,file="moto.rda")
```
